MCC Models

### 2019-08-06 papers/full_paper/SI_appendix.Rmd
finish full worked example

TWO CORRECTIONS:
- THE GENERATIVE PROCESS OF THE UTTERANCE: Speaker knows lexicon, but doesn't assume listener does, but must label (we do not model the decision as to whether or not to label). So the only thing informative they do is point.
- QUD is type, not referent. Change L0 model.

In this section, we work through a toy numerical example for how model predictions were generated for the pragmatics model. The prediction will correspond to the parameter free model described above (excluding the noise parameter). The values of the parameters are taken from the preference - same speaker - incongruent condition (see below); that is, the pressures of informativity and preference operate in conflicting directions. Fig. 1 shows a screenshot from the adult experiment, which the model was designed to capture. In this context, there are three potential referents of two types (pink-ish and yellow-ish; for simplicity, we refer refer the object's type by its color) on two tables: $r^{table: 1}_{type:pink}$, $r^{table: 2}_{type:pink}$; one yellow-ish, $r^{table: 2}_{type:yellow}$). For simplicity, we refer to the referents by the shorthand: $r_{p}^1, r_{p}^2, r_y^2$.
In principle, the speaker, the frog in this case, can produce one of four utterances, turning either to the left or right table and saying either label ("dax" or "wug"):  $u_1 = (\text{left}, dax)$, $u_2 = (\text{left}, wug)$, $u_3 = (\text{right}, dax)$, $u_4 = (\text{right}, wug)$.
We work out the example where the speaker produces utterance $u_3$, turning to the right table (two objects) and saying the label "dax" (though since the labels have no *a priori* meanings, the computation would be the same for utterance $u_4$).

@@

We assume that the intentional goal of the speaker is to get the listener to get an option of the correct type. That is, the informativity of an utterance is calculated with respect to conveying the correct object type as opposed to a particular referent (token).
First, we calculate the literal listener's posterior distribution over referents and lexica, and marginalize (average) over objects of the same type and the different lexica to compute the informativity of an utterance for a type.
The literal listener's posterior distribution over referents and lexica has support of size six, owing to the unique combinations of referents and lexica.
$$
\begin{aligned}
& P_{L_0}(r_p^1, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_p^1, u_3) P(r_1^r) P(L_1) = 0 \times 0.26 \times 0.5 =0\\
& P_{L_0}(r_p^2, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_p^2, u_3) P(r_2^r) P(L_1) = 1 \times 0.26 \times 0.5 =0.13\\
& P_{L_0}(r_y^2, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_y^2, u_3) P(r_3^y) P(L_1) = 0 \times 0.48 \times 0.5 =0\\
& P_{L_0}(r_p^1, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_p^1, u_3) P(r_1^r) P(L_2) = 0 \times 0.26 \times 0.5=0\\
& P_{L_0}(r_p^2, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_p^2, u_3) P(r_2^r) P(L_2) = 0 \times 0.26 \times 0.5=0\\
& P_{L_0}(r_y^2, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_y^2, u_3) P(r_3^y) P(L_2) = 1 \times 0.48 \times 0.5=0.24\\
\end{aligned}
$$

@@

Finally, based on equation 1 we can use these values to compute the probability that the listener thinks that the speaker is referring to the yellow object ($r_1$) when they produce $r_1$:

$$
\begin{aligned}
&P_{L_1}(t, \mathcal{L}|u ) = ...
\end{aligned}
$$

Thus, we normalize across all potential states that the utterance could refer to.

>>>>

In this section, we work through a toy numerical example for how model predictions were generated for the pragmatics model. The prediction will correspond to the parameter free model described above (excluding the noise parameter). The values of the parameters are taken from the preference - same speaker - congruent condition (see below); that is, the pressures of informativity and preference operate in the same direction. Fig. 1 shows a screenshot from the adult experiment, which the model was designed to capture. In this context, there are three potential referents of two types (pink-ish and yellow-ish; for simplicity, we refer refer the object's type by its color) on two tables: $r^{table: 1}_{type:pink}$, $r^{table: 2}_{type:pink}$; one yellow-ish, $r^{table: 2}_{type:yellow}$). For simplicity, we refer to the referents by the shorthand: $r_{p}^1, r_{p}^2, r_y^2$.
In principle, the speaker, the frog in this case, can produce one of four utterances, turning either to the left or right table and saying either label ("dax" or "wug"):  $u_1 = (\text{left}, dax)$, $u_2 = (\text{left}, wug)$, $u_3 = (\text{right}, dax)$, $u_4 = (\text{right}, wug)$.
We work out the example where the speaker produces utterance $u_3$, turning to the right table (two objects) and saying the label "dax" (though since the labels have no *a priori* meanings, the computation would be the same for utterance $u_4$).

@@

We assume that the intentional goal of the speaker is to get the listener to get an option of the correct type. That is, the informativity of an utterance is calculated with respect to conveying the correct object type as opposed to a particular referent (token).
First, we calculate the literal listener's posterior distribution over referents and lexica, and marginalize (average) over objects of the same type and the different lexica to compute the informativity of an utterance for a type.
The literal listener's posterior distribution over referents and lexica has support of size six, owing to the unique combinations of referents and lexica.
$$
\begin{aligned}
& P_{L_0}(r_p^1, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_p^1, u_3) P(r^1_r) P(\mathcal{L}_1) = 0 \times 0.26 \times 0.5 =0\\
& P_{L_0}(r_p^2, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_p^2, u_3) P(r^2_r) P(\mathcal{L}_1) = 1 \times 0.26 \times 0.5 =0.13\\
& P_{L_0}(r_y^2, \mathcal{L}_1 | {u_3})  \propto \mathcal{L}_1(r_y^2, u_3) P(r^2_y) P(\mathcal{L}_1) = 0 \times 0.48 \times 0.5 =0\\
& P_{L_0}(r_p^1, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_p^1, u_3) P(r^1_r) P(\mathcal{L}_2) = 0 \times 0.26 \times 0.5=0\\
& P_{L_0}(r_p^2, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_p^2, u_3) P(r^2_r) P(\mathcal{L}_2) = 0 \times 0.26 \times 0.5=0\\
& P_{L_0}(r_y^2, \mathcal{L}_2 | {u_3})  \propto \mathcal{L}_2(r_y^2, u_3) P(r^2_y) P(\mathcal{L}_2) = 1 \times 0.48 \times 0.5=0.24\\
\end{aligned}
$$

@@

Finally, based on equation 1 we can use these values to compute the probability that the listener thinks that the speaker is referring to the yellow type ($t_y$) when they produce $u_3$ (pointing to the right table and saying "dax"). In this case, it is the same as the probability of the yellow referent ($r^2_y$) because there is only one yellow object.

$$
\begin{aligned}
&P_{L_1}(r^2_y, \mathcal{L}|u ) = \frac{P_{S_1}(u_3 \mid t_y, \mathcal{L}) \cdot P(\mathcal{L}) \cdot P(r^2_y)}{
\sum_{r'} P_{S_1}(u_3 \mid r', \mathcal{L}) \cdot P(\mathcal{L}) \cdot P(r')
}
\end{aligned}
$$

$$
\begin{aligned}
& P_{L_1}(r_p^1, \mathcal{L}_1 | {u_3})  \propto P_{S_1}(u_3 \mid r_p^1, \mathcal{L}_1) P(r^1_r) P(\mathcal{L}_1) = 0.09 \times 0.26 \times 0.5 =0.012\\
& P_{L_1}(r_p^2, \mathcal{L}_1 | {u_3})  \propto P_{S_1}(u_3 \mid r_p^2, \mathcal{L}_1)  P(r^2_r) P(\mathcal{L}_1) = 0.09 \times 0.26 \times 0.5 =0.012\\
& P_{L_1}(r_y^2, \mathcal{L}_1 | {u_3})  \propto P_{S_1}(u_3 \mid r_y^2, \mathcal{L}_1)  P(r^2_y) P(\mathcal{L}_1) = 0 \times 0.48 \times 0.5 =0\\
& P_{L_1}(r_p^1, \mathcal{L}_2 | {u_3})  \propto P_{S_1}(u_3 \mid r_p^1, \mathcal{L}_2)  P(r^1_r) P(\mathcal{L}_2) = 0 \times 0.26 \times 0.5=0\\
& P_{L_1}(r_p^2, \mathcal{L}_2 | {u_3})  \propto P_{S_1}(u_3 \mid r_p^2, \mathcal{L}_2) P(r^2_r) P(\mathcal{L}_2) = 0 \times 0.26 \times 0.5=0\\
& P_{L_1}(r_y^2, \mathcal{L}_2 | {u_3})  \propto P_{S_1}(u_3 \mid r_y^2, \mathcal{L}_2) P(r^2_y) P(\mathcal{L}_2) = 1 \times 0.48 \times 0.5=0.24\\
\end{aligned}
$$

After normalizing, the full joint-posterior distribution over referents and lexica is:

$$
\begin{aligned}
& P_{L_1}(r_p^1, \mathcal{L}_1 | {u_3})  = 0.045\\
& P_{L_1}(r_p^2, \mathcal{L}_1 | {u_3})  = 0.045\\
& P_{L_1}(r_y^2, \mathcal{L}_1 | {u_3})  = 0 \\
& P_{L_1}(r_p^1, \mathcal{L}_2 | {u_3})  = 0\\
& P_{L_1}(r_p^2, \mathcal{L}_2 | {u_3})  = 0\\
& P_{L_1}(r_y^2, \mathcal{L}_2 | {u_3})  = 0.91\\
\end{aligned}
$$

To arrive at the distribution on types, you simply add all of the referents of the same type and marginalize out the lexicon:

$$
\begin{aligned}
&P_{L_1}(t_p|u_3) =  \sum_{i \in {1, 2}} P_{L_1}(r^1_p, \mathcal{L}_i|u_3) + P_{L_0}(r^2_p, \mathcal{L}_i|u_3) = 0.09 \\
&P_{L_1}(t_y|u_3) =  \sum_{i \in {1, 2}} P_{L_1}(r^2_y, \mathcal{L}_i|u_3)  = 0.91 \\
\end{aligned}
$$
